---
title: "3_Visual_types"
format: html
---

Think about how to transform the images because of larger differences in size between states 

This code creates all of the visuals for the states separated by state and type of visualization 
```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(purrr)

# Load in csv fule and make sure everything is ordered correctly and the correct data type 
visuals_data <- read_csv("Visuals_data_cleaned.csv",
                         col_types = cols(
                           State = col_character(),
                           Year = col_factor(levels = c("1860", "1870"), ordered = TRUE),
                           `Mentally Ill` = col_double(),
                           Blind = col_double(),
                           `Developmentally Disabled` = col_double(),
                           `Deaf-Mutes` = col_double()
                         ))

# Pivot longer to get category and count columns to more easily display the data 
visuals_long <- visuals_data %>%
  pivot_longer(cols = c(`Mentally Ill`, Blind, `Developmentally Disabled`, `Deaf-Mutes`),
               names_to = "Category",
               values_to = "Count")
 

# creates a folder to store the images in 
output_dir <- "plots_by_state_and_visual_type"
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

# Function to generate and save 3 plot types per state
create_and_save_plots <- function(state_name) {
  df <- visuals_long %>% filter(State == state_name)
  
  # Polar plot
  p_polar <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
    geom_col(position = "identity", alpha = 0.5) +
    scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
    # changes the scale to log values 
    scale_y_log10() +
    coord_polar(theta = "x") +
    facet_wrap(~Category, ncol = 2) +
    theme_void() +
    ggtitle(paste("Polar Plot -", state_name))
  
  ggsave(filename = file.path(output_dir, paste0(state_name, "-polar.png")),
         plot = p_polar, width = 7, height = 5, dpi = 300)
  
  # Overlaid bar plot
  p_overlaid <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
    geom_col(position = "identity", alpha = 0.5, color = "black") +
    scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      # changes the scale to log values 
    scale_y_log10() +
    facet_wrap(~Category, ncol = 2) +
    theme_bw() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.x = element_blank()) +
    ggtitle(paste("Overlaid Bar Plot -", state_name))
  
  ggsave(filename = file.path(output_dir, paste0(state_name, "-overlaid.png")),
         plot = p_overlaid, width = 7, height = 5, dpi = 300)
  
  # Dodged bar plot
  p_dodged <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
    geom_col(position = position_dodge2(), alpha = 0.5, color = "black") +
    scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      # changes the scale to log values 
    scale_y_log10() +
    facet_wrap(~Category, ncol = 2) +
    theme_bw() +
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.title.x = element_blank()) +
    ggtitle(paste("Dodged Bar Plot -", state_name))
  
  ggsave(filename = file.path(output_dir, paste0(state_name, "-dodged.png")),
         plot = p_dodged, width = 7, height = 5, dpi = 300)
}

# Loop through all states and generate plots
unique(visuals_long$State) %>% walk(create_and_save_plots)

```


This code separates by the 4 disability types and and 3 visual types
```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(purrr)
library(stringr)

# Load data
visuals_data_1 <- read_csv("Visuals_data_cleaned.csv",
                         col_types = cols(
                           State = col_character(),
                           Year = col_factor(levels = c("1860", "1870"), ordered = TRUE),
                           `Mentally Ill` = col_double(),
                           Blind = col_double(),
                           `Developmentally Disabled` = col_double(),
                           `Deaf-Mutes` = col_double()
                         ))

# Convert to long format
visuals_long <- visuals_data_1 %>%
  pivot_longer(cols = c(`Mentally Ill`, Blind, `Developmentally Disabled`, `Deaf-Mutes`),
               names_to = "Category",
               values_to = "Count") 

create_plot <- function(data, disability, vis_type) {
  df <- data %>% filter(Category == disability)
  
  if (vis_type == "polar") {
    # Facet by state to get one polar plot per state
    p <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
      geom_col(position = "identity", alpha = 0.5) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
        # changes the scale to log values 
    #scale_y_log10() +
      coord_polar(theta = "x") +
      facet_wrap(~ State, ncol = 6) +
      theme_void() +
      labs(title = paste("Polar Plot -", disability))
    
  } else {
    # Base for bar plots: all states on x axis
    base_plot <- ggplot(df, aes(x = State, y = Count, fill = Year)) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
     
        # changes the scale to sqrt values 
    #scale_y_sqrt() +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      labs(title = paste(vis_type, "-", disability),
           x = "State",
           y = "Count")
    
    if (vis_type == "overlaid") {
      p <- base_plot +
        geom_col(position = "identity", alpha = 0.5, color = "black") +
      
          # changes the scale to sqrt values 
    #scale_y_log10() +
        theme(axis.ticks.x = element_blank(),
              axis.title.x = element_blank())
      
    } else if (vis_type == "dodged") {
      p <- base_plot +
        geom_col(position = position_dodge(width = 0.7), alpha = 0.5, color = "black") +
     
          # changes the scale to sqrt values 
  #  scale_y_sqrt() +
        theme(axis.title.x = element_blank())
    } else {
      stop("Unknown visual type")
    }
  }
  
  return(p)
}

# Create output folder
if (!dir.exists("plots_by_disability_and_visual_type")) {
  dir.create("plots_by_disability_and_visual_type")
}

disabilities <- c("Mentally Ill", "Blind", "Developmentally Disabled", "Deaf-Mutes")
visual_types <- c("polar", "overlaid", "dodged")

walk(disabilities, function(dis) {
  walk(visual_types, function(vis) {
    file_name <- paste0("plots_by_disability_and_visual_type/", str_replace_all(dis, " ", "_"), "_", vis, ".png")
    plot_obj <- create_plot(visuals_long, dis, vis)
    ggsave(file_name, plot = plot_obj, width = 14, height = 7, dpi = 300)
  })
})

```




This creates visuals based on groupings made by sd of the data 

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(purrr)
library(stringr)

# Load data
visuals_data_1 <- read_csv("Visuals_data_cleaned.csv",
                          col_types = cols(
                            State = col_character(),
                            Year = col_factor(levels = c("1860", "1870"), ordered = TRUE),
                            `Mentally Ill` = col_double(),
                            Blind = col_double(),
                            `Developmentally Disabled` = col_double(),
                            `Deaf-Mutes` = col_double()
                          ))

# Convert to long format
visuals_long <- visuals_data_1 %>%
  pivot_longer(cols = c(`Mentally Ill`, Blind, `Developmentally Disabled`, `Deaf-Mutes`),
               names_to = "Category",
               values_to = "Count") 

# Calculate z-score and assign Group by Category + Year
grouped_data <- visuals_long %>%
  group_by(Category, Year) %>%
  mutate(
    z_score = (Count - mean(Count, na.rm = TRUE)) / sd(Count, na.rm = TRUE),
    Group = case_when(
      z_score < -1.5 ~ "Very Low",
      z_score >= -1.5 & z_score < -0.5 ~ "Low",
      z_score >= -0.5 & z_score < 0.5 ~ "Average",
      z_score >= 0.5 & z_score < 1.5 ~ "High",
      z_score >= 1.5 ~ "Very High",
     # TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

# Create wide format to get transition group per state & category
group_transitions <- grouped_data %>%
  select(State, Category, Year, Group) %>%
  pivot_wider(names_from = Year, values_from = Group, names_prefix = "Group_") %>%
  mutate(Transition = paste0(Group_1860, "-", Group_1870))

# Join transition info back to the long data for plotting
visuals_with_transitions <- visuals_long %>%
  left_join(
    group_transitions %>% select(State, Category, Transition),
    by = c("State", "Category")
  )


create_plot <- function(data, disability, transition_group, vis_type) {
  df <- data %>%
    filter(Category == disability, Transition == transition_group)
  
  if(nrow(df) == 0){
    message(glue::glue("No data for {disability} with transition {transition_group}"))
    return(NULL)
  }
  
  if (vis_type == "polar") {
    p <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
      geom_col(position = "identity", alpha = 0.5) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      coord_polar(theta = "x") +
      facet_wrap(~ State, ncol = 6) +
      theme_void() +
      labs(title = paste("Polar Plot -", disability, "\nTransition:", transition_group))
    
  } else {
    base_plot <- ggplot(df, aes(x = State, y = Count, fill = Year)) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      labs(title = paste(vis_type, "-", disability, "\nTransition:", transition_group),
           x = "State",
           y = "Count")
    
    if (vis_type == "overlaid") {
      p <- base_plot +
        geom_col(position = "identity", alpha = 0.5, color = "black") +
        theme(axis.ticks.x = element_blank(),
              axis.title.x = element_blank())
      
    } else if (vis_type == "dodged") {
      p <- base_plot +
        geom_col(position = position_dodge(width = 0.7), alpha = 0.5, color = "black") +
        theme(axis.title.x = element_blank())
    } else {
      stop("Unknown visual type")
    }
  }
  
  return(p)
}

# Create output folder
if (!dir.exists("grouped_vis")) {
  dir.create("grouped_vis")
}

disabilities <- unique(visuals_with_transitions$Category)
transitions <- unique(visuals_with_transitions$Transition)
visual_types <- c("polar", "overlaid", "dodged")

library(glue)

# Loop over each disability, each transition group, and each visual type
walk(disabilities, function(dis) {
  # Filter transitions for this disability (some transitions may not appear for each disability)
  trans_for_dis <- visuals_with_transitions %>%
    filter(Category == dis) %>%
    pull(Transition) %>%
    unique()
  
  walk(trans_for_dis, function(trans_grp) {
    walk(visual_types, function(vis) {
      file_name <- glue("grouped_vis/{str_replace_all(dis, ' ', '_')}_{str_replace_all(trans_grp, '-', '_')}_{vis}.png")
      plot_obj <- create_plot(visuals_with_transitions, dis, trans_grp, vis)
      
      if(!is.null(plot_obj)){
        ggsave(file_name, plot = plot_obj, width = 14, height = 7, dpi = 300)
      }
    })
  })
})

# Allows you to call these and actually assign visuals based upon this 
# Summarize each state's transitions across all categories
state_transition_summary <- group_transitions %>%
  mutate(Transition = paste0(Group_1860, "-", Group_1870)) %>%
  select(State, Category, Transition) %>%
  pivot_wider(names_from = Category, values_from = Transition) %>%
  arrange(State)

# Preview the result
head(state_transition_summary)

```

Other ideas: 
- make a filter for all the states that have negative growth 
- make a filter for all the states that fall within the JND(a few percent growth)


Goal: 
- find a few state comparisons from states that are similar sizes and see what ranges of true differences there are 





ACTUAL CODE BEING USED 

# pairing visuals 
```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(tidyr)
library(purrr)
library(stringr)

# Load data
visuals_data <- read_csv("Visuals_data_cleaned.csv",
                         col_types = cols(
                           State = col_character(),
                           Year = col_factor(levels = c("1860", "1870"), ordered = TRUE),
                           `Mentally Ill` = col_double(),
                           Blind = col_double(),
                           `Developmentally Disabled` = col_double(),
                           `Deaf-Mutes` = col_double()
                         ))

# Convert to long format
visuals_long <- visuals_data %>%
  pivot_longer(cols = c(`Mentally Ill`, Blind, `Developmentally Disabled`, `Deaf-Mutes`),
               names_to = "Category",
               values_to = "Count")

# Define state pairs - original mental illness pairs plus new developmentally disabled pairs
state_pairs <- list(
  # Mental illness pairs
  c("MISSISSIPPI", "LOUISIANA"),
  c("NORTH CAROLINA", "TENNESSEE"),
  c("ALABAMA", "GEORGIA"),
  c("FLORIDA", "DELAWARE"),
  c("MISSOURI", "OHIO"),
  # Developmentally disabled pairs
  c("MAINE", "KENTUCKY")
)

# Verify all states exist in data
all_states <- unique(visuals_long$State)
walk(state_pairs, function(pair) {
  if (!all(pair %in% all_states)) {
    missing <- setdiff(pair, all_states)
    stop(paste("States not found in data:", paste(missing, collapse = ", ")))
  }
})

# EXACT recreation of your original plotting function style
create_pair_plot <- function(data, disability, vis_type, states) {
  df <- data %>% 
    filter(Category == disability, State %in% states) %>%
    mutate(State = factor(State, levels = states))  # Preserve order
  
  if (nrow(df) == 0) {
    stop(paste("No data found for:", paste(states, collapse = " vs ")))
  }
  
  if (vis_type == "polar") {
    p <- ggplot(df, aes(x = 1, y = Count, fill = Year)) +
      geom_col(position = "identity", alpha = 0.5) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      coord_polar(theta = "x") +
      facet_wrap(~ State, ncol = 2) +
      theme_void() +
      labs(title = paste("Polar Plot -", disability))
    
  } else {
    # EXACT original bar plot styling
    base_plot <- ggplot(df, aes(x = State, y = Count, fill = Year)) +
      scale_fill_manual(values = c("1860" = "grey90", "1870" = "black")) +
      theme_bw() +  # Original used theme_bw()
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      labs(title = paste(vis_type, "-", disability),
           x = "State",
           y = "Count")
    
    if (vis_type == "overlaid") {
      p <- base_plot +
        geom_col(position = "identity", alpha = 0.5, color = "black") +
        theme(axis.ticks.x = element_blank(),
              axis.title.x = element_blank())
      
    } else if (vis_type == "dodged") {
      p <- base_plot +
        # width changes how overlaid they are. 1 is not overlaid and the closer to 0 the more overlaid
        geom_col(position = position_dodge(width = .8), width = 0.7, alpha = 0.5, color = "black") +
        theme(axis.title.x = element_blank())
    } else {
      stop("Unknown visual type")
    }
  }
  
  return(p)
}

# Create output folder
if (!dir.exists("pair_plots_original_style")) {
  dir.create("pair_plots_original_style")
}

# Define visualization types
visual_types <- c("polar", "overlaid", "dodged")

# Generate all plots for each pair
walk(state_pairs[1:5], function(pair) {  # Original mental illness pairs
  walk(visual_types, function(vis) {
    tryCatch({
      file_name <- paste0("pair_plots_original_style/Mentally_Ill_", 
                         paste(pair, collapse = "_"), "_", 
                         vis, ".png")
      plot_obj <- create_pair_plot(visuals_long, "Mentally Ill", vis, pair)
      ggsave(file_name, plot = plot_obj, width = 6, height = 4, dpi = 300)
    }, error = function(e) {
      message(paste("Error creating plot for", paste(pair, collapse = " vs "), ":", e$message))
    })
  })
})

# Generate developmentally disabled plots for Maine and Kentucky
walk(state_pairs[6], function(pair) {  # Maine vs Kentucky
  walk(visual_types, function(vis) {
    tryCatch({
      file_name <- paste0("pair_plots_original_style/Developmentally_Disabled_", 
                         paste(pair, collapse = "_"), "_", 
                         vis, ".png")
      plot_obj <- create_pair_plot(visuals_long, "Developmentally Disabled", vis, pair)
      ggsave(file_name, plot = plot_obj, width = 6, height = 4, dpi = 300)
    }, error = function(e) {
      message(paste("Error creating plot for", paste(pair, collapse = " vs "), ":", e$message))
    })
  })
})

message("Plot generation complete. Check 'pair_plots_original_style' folder.")
```


# creating the example plot 
```{r}
library(ggplot2)

# Create the data
plot_data <- data.frame(
  Year = factor(c("1860", "1870")),
  Count = c(130, 200),
  State = "State 1"
)

# Create the dodged bar plot with your exact color scheme
dodged_plot <- ggplot(plot_data, aes(x = State, y = Count, fill = Year)) +
  geom_col(
    position = position_dodge(width = 0.8), 
    width = 0.7,
    color = "black",        # Black outlines
    alpha = 0.5            # Slight transparency
  ) +
  scale_fill_manual(
    values = c(
      "1860" = "grey90",   # Light gray for 1860
      "1870" = "black"     # Black for 1870
    )
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  ) +
  labs(
    title = "Example Bar Chart",
    x = "",
    y = "Count",
    fill = "Year"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

# Print the plot
print(dodged_plot)

# Save the plot (optional)
ggsave("pair_plots_original_style/dodged_example.png", plot = dodged_plot, width = 6, height = 3, dpi = 300)
```

